Effectful Semantics for Substructural SSA

### 2023-11-04 pldi24/paper.tex
Elgot category properties
Began work on refactoring

% Syntax
\newcommand{\letexpr}[3]{\ensuremath{\ms{let}\;#1 = #2\;\ms{in}\;#3}}
\newcommand{\letstmt}[3]{\ensuremath{\ms{let}\;#1 = #2; #3}}
\newcommand{\brb}[2]{\ms{br}\;#1\;#2}
\newcommand{\lbrb}[2]{\brb{\lbl{#1}}{#2}}
\newcommand{\ite}[3]{\ms{if}\;#1\;\{#2\}\;\ms{else}\;\{#3\}}
\newcommand{\ewhere}[2]{\ms{then}\;#1\;\ms{where}\;#2}
\newcommand{\where}[2]{#1\;\ms{where}\;#2}
\newcommand{\wbranch}[3]{#1(#2) \lto #3}
\newcommand{\lwbranch}[3]{\wbranch{\lbl{#1}}{#2}{#3}}
\newcommand{\bsplice}[3]{#1(#2)\;\{#3\}}
\newcommand{\lbsplice}[3]{\bsplice{\lbl{#1}}{#2}{#3}}
\newcommand{\csplits}[3]{#1 \mapsto #2;#3}
\newcommand{\cwk}[2]{#1 \mapsto #2}
\newcommand{\lwk}[2]{#1 \rightsquigarrow #2}
\newcommand{\tlin}[2]{#2 \subseteq \ms{lin}(#1)}
\newcommand{\ltlin}[3]{#3 \subseteq \ms{lin}(#1) \cap #2}
\newcommand{\thyp}[3]{#1: {#2}^{#3}}
\newcommand{\lhyp}[3]{#1[#2](#3)}
\newcommand{\llhyp}[3]{\lhyp{\lbl{#1}}{#2}{#3}}
\newcommand{\rle}[1]{{\scriptsize\textsf{#1}}}
\newcommand{\taff}{\ms{a}}
\newcommand{\trel}{\ms{r}}
\newcommand{\tint}{\infty}
\newcommand{\hasty}[5]{#1 \vdash_{#2} #3: {#4}^{#5}}
\newcommand{\haslb}[3]{#1 \vdash #2 \rhd #3}
\newcommand{\lhaslb}[3]{#1 \vdash #2 \rhd #3}
\newcommand{\issubst}[3]{#1: #2 \mapsto #3}
\newcommand{\lbsubst}[3]{#1: #2 \rightsquigarrow #3}
\newcommand{\exprletsubst}[2]{{#1};{#2}}
\newcommand{\stmtletsubst}[2]{{#1};{#2}}
\newcommand{\mhole}[1]{{#1}^?}
\newcommand{\lhole}[1]{?#1}
\newcommand{\mhasty}[6]{#1;#2 \vdash_{#3} #4: {#5}^{#6}}
\newcommand{\mhaslb}[4]{#1;#2 \vdash #3 \rhd #4}
\newcommand{\mlhaslb}[4]{#1;#2 \vdash #3 \rhd #4}
\newcommand{\tyhole}[5]{#1: #2 \mapsto_{#3} {#4}^{#5}}
\newcommand{\blkhole}[3]{#1: #2 \mapsto #3}
\newcommand{\cfghole}[3]{#1: #2 \mapsto #3}
\newcommand{\substctx}[2]{{#1}^{#2}}
\newcommand{\substlbs}[2]{{#1}^{#2}}
\newcommand{\restrictsubst}[2]{{#1}_{#2}}
\newcommand{\subsubst}[2]{#1 \subseteq #2}
\newcommand{\isrw}[3]{#1: #2 \mapsto #3}
% \newcommand{\strictlbsubst}[3]{#1: #2 \rightsquigarrow_= #3}

@@

\TODO{define premonoidal category}

@@

\begin{definition}[Substructural Monoidal Category]
  A \textbf{substructural monoidal category} is a tuple \((\mc{C}, \mc{A}, \mc{R})\) where:
  \begin{itemize}
    \item \(\mc{C}\) is a family of categories \(\mc{C}^q\) parametrized by subsets \(q \subseteq \{\ms{a}, \ms{r}\}\) such that
    \begin{itemize}
      \item For all \(q \subseteq q'\), \(\mc{C}^{q'}\) is a wide subcategory of \(\mc{C}^q\), and in particular all object sets \(|\mc{C}^q|\) are the same. We will write this set as \(|\mc{C}|\)
      \item All \(\mc{C}^q\) share symmetric monoidal structure, and are individually closed under the tensor products of morphisms
      \item \(\mc{C}^{q \cap q'} = \mc{C}^q \cap \mc{C}^{q'}\)
    \end{itemize}
    \item \(\mc{A} \subseteq |\mc{C}|\) is a set of \textbf{affine} objects closed under tensor products, such that
    \begin{itemize}
      \item \(\mb{1} \in \mc{A}\)
      \item Every \(A\) in \(\mc{A}\) is equipped with a morphism \(\ms{drop}(A): \mc{C}^{\{\ms{a}, \ms{r}\}}(A, \mb{1})\)
      \item \(\forall A, B \in \mc{A}, \ms{drop}(A \otimes B) = \ms{drop}(A) \otimes \ms{drop}(B);\lambda_I\)
      \item \(\forall A, B \in \mc{A}, \forall f: \mc{C}^{\{\ms{a}\}}(A, B), f;\ms{drop}(B) = \ms{drop}(A)\)
    \end{itemize}
    \item \(\mc{R} \subseteq |\mc{C}|\) is a set of \textbf{relevant} objects closed under tensor products, such that
    \begin{itemize}
      \item \(\mb{1} \in \mc{R}\)
      \item Every \(A\) in \(\mc{R}\) is equipped with a morphism \(\ms{split}(A): \mc{C}^{\{\ms{a}, \ms{r}\}}(A, A \otimes A)\)
      \item \(\forall A, B \in \mc{R}, \ms{split}(A \otimes B) = \ms{split}(A) \otimes \ms{split}(B);\alpha;A \otimes \sigma_{A, B} \otimes B;\alpha\)
      \item Associativity: \(\forall A \in \mc{R}, \ms{split}(A);\ms{split}(A) \otimes A;\alpha = \ms{split}(A);A \otimes \ms{split}(A)\)
      \item Commutativity: \(\forall A \in \mc{R}, \ms{split}(A);\sigma = \ms{split}(A)\)
      \item Comonoid: \(\forall A \in \mc{A} \cap \mc{R}, \ms{split}(A);A \otimes \ms{drop}(A) = \rho_A^{-1} \quad \ms{split}(A);\ms{drop}(A) \otimes A = \lambda_A^{-1}\)
    \end{itemize}
  \end{itemize}
  We call the morphisms in \(\mc{C}^{\{\ms{a}\}}\) \textbf{affine}, the morphisms in \(\mc{C}^{\{\ms{r}\}}\) \textbf{relevant}, and the morphisms in \(\mc{C}^{\{\ms{a}, \ms{r}\}}\) \textbf{pure} or \textbf{intuitionistic}. We call the morphisms in \(\mc{C}^\varnothing\) but not necessarily in any of the other \(\mc{C}^q\) \textbf{linear}. We call a substructural monoidal category a \textbf{substructural Cartesian category} if \(\mc{A} = \mc{R} = |\mc{C}|\), in which case \(\mc{C}^{\{\ms{a}, \ms{r}\}}\) is Cartesian.
\end{definition}
A simple example of a substructural Cartesian category is the category of relations \(\ms{Rel}\), viewed as the Kleisli category of the power-set monad with morphisms of the form \(\ms{Rel}(A, B) = A \to \mc{P}(B)\). In particular, we can define:
\begin{equation}
  \begin{aligned}
    \ms{Rel}^\varnothing(A, B) 
    &= \ms{Rel}(A, B) 
    &&= A \to \mc{P}(B) \\
    \ms{Rel}^{\{\ms{a}\}}(A, B) 
    &= \{R \in \ms{Rel}(A, B) \mid \forall a, |R(a)| \geq 1\} 
    &&= A \to \mc{P}^+(B) \\
    \ms{Rel}^{\{\ms{r}\}}(A, B) 
    &= \{R \in \ms{Rel}(A, B) \mid \forall a, |R(a)| \leq 1\} 
    &&= A \to \ms{Option}(B) \\
    \implies \ms{Rel}^{\{\ms{a}, \ms{r}\}}(A, B) 
    &= \{R \in \ms{Rel}(A, B) \mid \forall a, |R(a)| = 1\} 
    &&= A \to B
  \end{aligned}
\end{equation}
Here, \(\ms{split}(A)\) is just the copy operator \(a \mapsto (a, a)\) and \(\ms{drop}(A)\) the deletion operator \(a \mapsto ()\); it is quite trivial to prove that these lie in \(\ms{Rel}^{\{\ms{a}, \ms{r}\}}\), being functions, and that they satisfy the axioms of a commutative comonoid, as desired.

@@

\begin{definition}[Elgot category] 
  An \textbf{Elgot category} is a category \(\mc{C}\) with coproducts equipped with an operator \(\cdot^\dagger: \mc{C}(A, B + A) \to \mc{C}(A, B)\) satisfying the following properties:
  \begin{itemize}
    \item \textbf{Fixpoint:} ...
    \item \textbf{Naturality:} ...
    \item \textbf{Codiagonal:} ...
    \item \textbf{Uniformity:} ...
  \end{itemize}
  A monad \(\mb{T}\) on \(\mc{C}\) such that its Kleisli cateogry \(\mc{C}_{\mb{T}}\) is Elgot is called an \textbf{Elgot monad}.
\end{definition}

\TODO{distributivity of monads}

@@

\TODO{refactor below}

The semantics for an \isotopessa dialect are given by:
\begin{itemize}
  \item A family of categories \(\mc{C}^q_1\) parametrized by quantity \(q \subseteq \{\trel, \taff\}\)
  \item \(q \subseteq q'\) implies \(\mc{C}^{q'}_1\) is a wide subcategory of \(\mc{C}^q_1\)
  \item All \(\mc{C}^q_1\) share symmetric monoidal structure (i.e., the inclusion functors between them are strict symmetric monoidal functors).
  \item A symmetric premonoidal category \(\mc{C}_0\) equipped with an Elgot structure
  \item An object \(\mb{2}\), morphisms \(\ms{tt}, \ms{ff}: \mc{C}_1^\tint(\mb{1}, \mb{2})\), and a family of isomorphisms \(\ms{ite}_A: 2 \otimes A \simeq A + A\) such that 
  \begin{itemize}
    \item \(\ms{ite}_A^{-1} = [\lambda_A^{-1};\ms{tt} \otimes A, \lambda_A^{-1};\ms{ff} \otimes A]\)
    \item \(\mb{2} \otimes f;\ms{ite}_A = \ms{ite}_A;f + f\)
  \end{itemize}
  \item An identity on objects functor functor \(\upg{\cdot}{}: \mc{C}^\varnothing_1 \to \mc{C}_0\) which preserves all symmetric premonoidal structure
  \item A mapping from base types \(X\) to objects in \(\dnt{X}: |\mc{C}|\)
  \item A mapping from functions \(f \in \mc{I}_p^q(A, B)\) to morphisms \(\dnt{f}: \mc{C}_p^q(\dnt{A}, \dnt{B})\), where \(\mc{C}_0^q = \mc{C}_0\)
  \item A set of \textit{affine} objects \(A, B \in \ms{Aff}(\mc{C}) \subseteq |\mc{C}|\) equipped with maps \(\ms{drop}(A): \mc{C}_1^\tint(A, \mb{1})\) such that:
  \begin{itemize}
    \item \(\ms{drop}(A \otimes B) = \ms{drop}(A) \otimes \ms{drop}(B); \lambda_\mb{1}\)
    \item For all \(f: \mc{C}^\taff_1(A, B), f;\ms{drop}(B) = \ms{drop}(A)\)
  \end{itemize}
  \item A set of \textit{relevant} objects \(A, B \in \ms{Rel}(\mc{C}) \subseteq |\mc{C}|\) equipped with maps \(\ms{split}(A): \mc{C}_1^\tint(A, A \times A)\) such that:
  \begin{itemize}
    \item \(\ms{split}(A \otimes B) = \ms{split}(A) \otimes \ms{split}(B);\alpha;A \otimes \sigma_{A, B} \otimes B;\alpha\)
    \item For all \(f: \mc{C}^\trel_1(A, B), f;\ms{split}(B) = \ms{split}(A);f \otimes f\)
    \item \textit{Commutativity}: \(\ms{split}(A);\sigma = \ms{split}(A)\)
    \item \textit{Associativity}: \(\ms{split}(A);A \otimes \ms{split}(A) = \ms{split}(A);\ms{split}(A) \otimes A;\alpha\)
    \item \textit{Comonoid}: if \(A \in \ms{Aff}(\mc{C})\), \(\ms{split}(A);\ms{drop}(A) \otimes A = \lambda^{-1}\) and \(\ms{split}(A);A \otimes \ms{drop}(A) = \rho^{-1}\)
  \end{itemize}
\end{itemize}

>>>>

% Syntax
\newcommand{\letexpr}[3]{\ensuremath{\ms{let}\;#1 = #2\;\ms{in}\;#3}}
\newcommand{\letstmt}[3]{\ensuremath{\ms{let}\;#1 = #2; #3}}
\newcommand{\brb}[2]{\ms{br}\;#1\;#2}
\newcommand{\lbrb}[2]{\brb{\lbl{#1}}{#2}}
\newcommand{\ite}[3]{\ms{if}\;#1\;\{#2\}\;\ms{else}\;\{#3\}}
\newcommand{\ewhere}[2]{\ms{then}\;#1\;\ms{where}\;#2}
\newcommand{\where}[2]{#1\;\ms{where}\;#2}
\newcommand{\wbranch}[3]{#1(#2) \lto #3}
\newcommand{\lwbranch}[3]{\wbranch{\lbl{#1}}{#2}{#3}}
\newcommand{\bsplice}[3]{#1(#2)\;\{#3\}}
\newcommand{\lbsplice}[3]{\bsplice{\lbl{#1}}{#2}{#3}}
\newcommand{\csplits}[3]{#1 \mapsto #2;#3}
\newcommand{\cwk}[2]{#1 \mapsto #2}
\newcommand{\lwk}[2]{#1 \rightsquigarrow #2}
\newcommand{\tlin}[2]{#2 \subseteq \ms{lin}(#1)}
\newcommand{\ltlin}[3]{#3 \subseteq \ms{lin}(#1) \cap #2}
\newcommand{\thyp}[3]{#1: {#2}^{#3}}
\newcommand{\lhyp}[3]{#1[#2](#3)}
\newcommand{\llhyp}[3]{\lhyp{\lbl{#1}}{#2}{#3}}
\newcommand{\rle}[1]{{\scriptsize\textsf{#1}}}
\newcommand{\taff}{\ms{a}}
\newcommand{\trel}{\ms{r}}
\newcommand{\tint}{\{\ms{a}, \ms{r}\}}
\newcommand{\hasty}[5]{#1 \vdash_{#2} #3: {#4}^{#5}}
\newcommand{\haslb}[3]{#1 \vdash #2 \rhd #3}
\newcommand{\lhaslb}[3]{#1 \vdash #2 \rhd #3}
\newcommand{\issubst}[3]{#1: #2 \mapsto #3}
\newcommand{\lbsubst}[3]{#1: #2 \rightsquigarrow #3}
\newcommand{\exprletsubst}[2]{{#1};{#2}}
\newcommand{\stmtletsubst}[2]{{#1};{#2}}
\newcommand{\mhole}[1]{{#1}^?}
\newcommand{\lhole}[1]{?#1}
\newcommand{\mhasty}[6]{#1;#2 \vdash_{#3} #4: {#5}^{#6}}
\newcommand{\mhaslb}[4]{#1;#2 \vdash #3 \rhd #4}
\newcommand{\mlhaslb}[4]{#1;#2 \vdash #3 \rhd #4}
\newcommand{\tyhole}[5]{#1: #2 \mapsto_{#3} {#4}^{#5}}
\newcommand{\blkhole}[3]{#1: #2 \mapsto #3}
\newcommand{\cfghole}[3]{#1: #2 \mapsto #3}
\newcommand{\substctx}[2]{{#1}^{#2}}
\newcommand{\substlbs}[2]{{#1}^{#2}}
\newcommand{\restrictsubst}[2]{{#1}_{#2}}
\newcommand{\subsubst}[2]{#1 \subseteq #2}
\newcommand{\isrw}[3]{#1: #2 \mapsto #3}
% \newcommand{\strictlbsubst}[3]{#1: #2 \rightsquigarrow_= #3}

@@

\TODO{define premonoidal category?}

@@

\TODO{substructural premonoidal category, relationship? This makes \(\mc{C}_0^q\) have structure, which is useful at least for things that are affine but not central (see: dead code elimination). Duplication is Less Useful (TM) but nice for symmetry.}

\begin{definition}[Substructural Monoidal Category]
  A \textbf{substructural monoidal category} is a tuple \((\mc{C}, \mc{A}, \mc{R})\) where:
  \begin{itemize}
    \item \(\mc{C}\) is a family of categories \(\mc{C}^q\) parametrized by subsets \(q \subseteq \{\ms{a}, \ms{r}\}\) such that
    \begin{itemize}
      \item For all \(q \subseteq q'\), \(\mc{C}^{q'}\) is a wide subcategory of \(\mc{C}^q\), and in particular all object sets \(|\mc{C}^q|\) are the same. We will write this set as \(|\mc{C}|\)
      \item All \(\mc{C}^q\) share symmetric monoidal structure, and are individually closed under the tensor products of morphisms
      \item \(\mc{C}^{q \cap q'} = \mc{C}^q \cap \mc{C}^{q'}\)
    \end{itemize}
    \item \(\mc{A} \subseteq |\mc{C}|\) is a set of \textbf{affine} objects closed under tensor products, such that
    \begin{itemize}
      \item \(\mb{1} \in \mc{A}\)
      \item Every \(A\) in \(\mc{A}\) is equipped with a morphism \(\ms{drop}(A): \mc{C}^{\{\ms{a}, \ms{r}\}}(A, \mb{1})\)
      \item \(\forall A, B \in \mc{A}, \ms{drop}(A \otimes B) = \ms{drop}(A) \otimes \ms{drop}(B);\lambda_I\)
      \item \(\forall A, B \in \mc{A}, \forall f: \mc{C}^{\{\ms{a}\}}(A, B), f;\ms{drop}(B) = \ms{drop}(A)\)
    \end{itemize}
    \item \(\mc{R} \subseteq |\mc{C}|\) is a set of \textbf{relevant} objects closed under tensor products, such that
    \begin{itemize}
      \item \(\mb{1} \in \mc{R}\)
      \item Every \(A\) in \(\mc{R}\) is equipped with a morphism \(\ms{split}(A): \mc{C}^{\{\ms{a}, \ms{r}\}}(A, A \otimes A)\)
      \item \(\forall A, B \in \mc{R}, \ms{split}(A \otimes B) = \ms{split}(A) \otimes \ms{split}(B);\alpha;A \otimes \sigma_{A, B} \otimes B;\alpha\)
      \item Associativity: \(\forall A \in \mc{R}, \ms{split}(A);\ms{split}(A) \otimes A;\alpha = \ms{split}(A);A \otimes \ms{split}(A)\)
      \item Commutativity: \(\forall A \in \mc{R}, \ms{split}(A);\sigma = \ms{split}(A)\)
      \item Comonoid: \(\forall A \in \mc{A} \cap \mc{R}, \ms{split}(A);A \otimes \ms{drop}(A) = \rho_A^{-1} \quad \ms{split}(A);\ms{drop}(A) \otimes A = \lambda_A^{-1}\)
    \end{itemize}
  \end{itemize}
  We call the morphisms in \(\mc{C}^{\{\ms{a}\}}\) \textbf{affine}, the morphisms in \(\mc{C}^{\{\ms{r}\}}\) \textbf{relevant}, and the morphisms in \(\mc{C}^{\{\ms{a}, \ms{r}\}}\) \textbf{pure} or \textbf{intuitionistic}. We call the morphisms in \(\mc{C}^\varnothing\) but not necessarily in any of the other \(\mc{C}^q\) \textbf{linear}. We call a substructural monoidal category a \textbf{substructural Cartesian category} if \(\mc{A} = \mc{R} = |\mc{C}|\), in which case \(\mc{C}^{\{\ms{a}, \ms{r}\}}\) is Cartesian.
\end{definition}
\TODO{Define \(\mc{C}^{\ms{a}}, \mc{C}^{\ms{r}}\) syntax? Make lattice generated by syntax?}
A simple example of a substructural Cartesian category is the category of relations \(\ms{Rel}\), viewed as the Kleisli category of the power-set monad with morphisms of the form \(\ms{Rel}(A, B) = A \to \mc{P}(B)\). In particular, we can define:
\begin{equation}
  \begin{aligned}
    \ms{Rel}^\varnothing(A, B) 
    &= \ms{Rel}(A, B) 
    &&= A \to \mc{P}(B) \\
    \ms{Rel}^{\{\ms{a}\}}(A, B) 
    &= \{R \in \ms{Rel}(A, B) \mid \forall a, |R(a)| \geq 1\} 
    &&= A \to \mc{P}^+(B) \\
    \ms{Rel}^{\{\ms{r}\}}(A, B) 
    &= \{R \in \ms{Rel}(A, B) \mid \forall a, |R(a)| \leq 1\} 
    &&= A \to \ms{Option}(B) \\
    \implies \ms{Rel}^{\{\ms{a}, \ms{r}\}}(A, B) 
    &= \{R \in \ms{Rel}(A, B) \mid \forall a, |R(a)| = 1\} 
    &&= A \to B
  \end{aligned}
\end{equation}
Here, \(\ms{split}(A)\) is just the copy operator \(a \mapsto (a, a)\) and \(\ms{drop}(A)\) the deletion operator \(a \mapsto ()\); it is quite trivial to prove that these lie in \(\ms{Rel}^{\{\ms{a}, \ms{r}\}}\), being functions, and that they satisfy the axioms of a commutative comonoid, as desired.

@@

\begin{definition}[Elgot category] 
  An \textbf{Elgot category} is a category \(\mc{C}\) with coproducts equipped with an operator \(\cdot^\dagger: \mc{C}(A, B + A) \to \mc{C}(A, B)\) satisfying the following properties:
  \begin{itemize}
    \item \textbf{Fixpoint:} \(f;[\ms{id}, f^\dagger] = f^\dagger\)
    \item \textbf{Naturality:} \((f;g + \ms{id})^\dagger = f^\dagger;g\)
    \item \textbf{Codiagonal:} \(\forall f \in \mc{C}(A, (B + A) + A), (f^\dagger)^\dagger = (f;[\ms{id}, \iota_1])^\dagger\)
    \item \textbf{Uniformity:}
    \(
      h;f = g;\ms{id} + h \implies h;f^\dagger = g^\dagger
    \)
  \end{itemize}
  A monad \(\mb{T}\) on \(\mc{C}\) such that its Kleisli cateogry \(\mc{C}_{\mb{T}}\) is Elgot is called an \textbf{Elgot monad}.
\end{definition}
\TODO{why do these matter? they show up in the proofs, but...}
Some important properties of all Elgot categories include:
\begin{itemize}
  \item \textbf{Dinaturality:} \((g;[\iota_0, h])^\dagger = g;[\ms{id}, (h;[\iota_0, g])^\dagger]\). \TODO{Proposition 16 of \cite{goncharov-guarded-unguarded}; also Lemma 31 of \cite{goncharov-squaring}}
  \item \textbf{Squaring:} \((f;[\iota_0, f])^\dagger = f^\dagger\). \TODO{Lemma 31 of \cite{goncharov-squaring}}
\end{itemize}
\TODO{Hyland and Hasegawa traces?}
\TODO{intuition? Can draw diagrams, but those take up space}

\TODO{results on distributivity of monads}

@@

The semantics for an \isotopessa dialect with a set of types \(\mc{T}\) and instructions \(\mc{I}\) are given by:
\begin{itemize}
  \item A \textit{substructural effectful category} \(\mc{C}\) such that \(\mc{C}_0\) is \textit{distributive} and \textit{Elgot}, equipped with
  \item A function \(\ms{base}(X)\) from base types \(X \in \mc{T}\) to objects in \(|\mc{C}|\), such that \(\ms{a} \in \ms{lin}(X) \implies X \in \ms{Aff}(\mc{C})\) and \(\ms{r} \in \ms{lin}(X) \implies X \in \ms{Rel}(\mc{C})\)
  \item A function \(\ms{inst}(f)\) from instructions \(f \in \mc{I}_p^q(A, B)\) to morphisms in \(\mc{C}_p^q(\dnt{A}, \dnt{B})\)
\end{itemize}
\TODO{make aff and rel syntax nicer/more consistent here? Subcategory syntax would be nice for lemmas and just coherence.}
